---
title: "EDLD Final Project v2"
author: "Frances Pinkham, Eunji Kong, Elyse Calhoun, Yitong Jiang"
format: html
editor: visual
embed-resources: true
code-fold: true
---

```{r}
#| include: false
library(tidyverse)
library(here)
library(rio)
library(modelsummary)
library(knitr)
library(gt)
```

```{r}
schooling <- import(here("data","Schooling.csv"))
```

```{r}
schoolingsub <- schooling %>%
  select(rownames,ed76, daded, momed, wage76, iqscore, age76, nearc2, nearc4)

```

```{r}
schoolingsub1 <- schoolingsub |> 
  mutate(near_college = ifelse(nearc2 == "yes" & nearc4 =="yes", "near_colleges", 
                               ifelse(nearc2 == "yes" & nearc4 =="no","near_cc", 
                                      ifelse(nearc2 == "no" & nearc4 =="yes","near_4yr", "not_near_college")))) 
```

```{r}
schoolingsub1 <- schoolingsub1 %>%
  pivot_longer(
    cols = c(daded,momed),
    names_to = "parent",
    values_to = "par_education")

schoolingsub1$parent <- gsub("daded", "dad", schoolingsub1$parent)

schoolingsub1$parent <- gsub("momed", "mom", schoolingsub1$parent)
```

### Table 1: Demographics

```{r}
schoolingsub1 %>% 
    pivot_wider (names_from = parent,
                values_from = par_education) %>%
   select(-near_college) %>%
  gt_preview(schoolingsub1, top_n = 20, bottom_n = 1) %>%
  cols_label(ed76 = "Participant Education (Yrs)",
             wage76 = "Participant Wage (cents/hr)",
             iqscore = "IQ Score",
             age76 = "Age of Participant",
             nearc2 = "Near Community College",
             nearc4 = "Near Four-Year College",
             dad = "Paternal Education (Yrs)",
             mom = "Maternal Education (Yrs)",
             rownames = "Participant ID")
  
```

Research question 1: What is the relationship between parent education and participant's education?

### Plot 1

```{r}
schoolingsub1 |> 
  ggplot(aes(par_education, ed76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + theme_minimal() +
  labs(title = "Relationship between Parent and Participant Education",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Education (yrs)",
       color = "Parent"
       )

```

Research question 2: What is the relationship between parent education and participant's wage?

### Plot 2

```{r}
schoolingsub1 |> 
  ggplot(aes(par_education, wage76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + theme_minimal() +
   labs(title = "Relationship between Parent Education and Participant Wages",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Wages (cents/hr)",
       color = "Parent"
       )
```

Research question 3: Is there a difference in wage depending on whether you lived near a college?

<<<<<<< Updated upstream
### Table 2: Proximity and Wages
=======
### Table 1
>>>>>>> Stashed changes

```{r}
prox_wage <- schoolingsub1 %>%
  mutate(near_college = fct_recode(near_college, "Near 4-year College" ="near_4yr", 
                                   "Near Community College" = "near_cc",
                                   "Near 4-year and Community College" = "near_colleges",
                                   "Not Near College" = "not_near_college")) %>%
group_by(near_college) %>%
summarize(avg_wage = mean(wage76, na.rm = TRUE)/100) %>%
rename("Proximity to College" = near_college, "Participant Average Wage per Hour" = avg_wage) %>%

DT::datatable()
```

# Results

**Table 1** outlines our basic demographics for our study. For our participant demographics, we examined the education in years, IQ score, age, and proximity to higher education institutions. We also examined the level of maternal and paternal education level in years.

<<<<<<< Updated upstream
**Plot 1** reflects a linear regression model of participant education as a function of parental education. Although we have not performed a complete analysis on these descriptive statistics, our model may indicate a positive correlation between parental education and participant education. Both maternal and paternal education levels may be positively correlated with participant education levels.

**Plot 2** also indicates a possible positive correlation between parental education and participant wages, shown in cents/hour. Further statistical testing would be required to verify this.

```{r}
#| include: false
min_wage <- 2.30
near_fouryear <- 6.45
```

**Table 2** displays the participant's proximity to a college alongside their average wage per hour. Those participants who lived near both a four-year and a community college had the highest average wage per hour, at \$6.45. According to the [Department of Labor](https://www.dol.gov/agencies/whd/minimum-wage/history), the minimum wage in the USA in 1976 was \$2.30 per hour, which means that these participants' wages were `r near_fouryear/min_wage` times higher than the minimum wage.

Those who resided near only a 4-year college had slightly lower average wages, at approximately \$5.68/hr.

Those who were not near any college had marginally higher average wages (\$5.18/hour) than those near a community college, at \$5.13 per hour.
=======
schoolingsub1 %>%   
  group_by(near_college) |> 
  summarize(avg_wage = mean(wage76, na.rm = TRUE)/100) |>
  rename("Proximity to College" = near_college, "Participant Average Wage per Hour" = avg_wage) %>%
DT::datatable()
```
>>>>>>> Stashed changes
