---
title: "EDLD Final Project v2"
author: "Frances Pinkham, Eunji Kong, Elyse Calhoun, Yitong Jiang"
format: html
editor: visual
embed-resources: true
code-fold: true
---

```{r}
#| include: false
library(tidyverse)
library(here)
library(rio)
library(modelsummary)
```

```{r}
schooling <- import(here("data","Schooling.csv"))
```

```{r}
schoolingsub <- schooling %>%
  select(rownames,ed76, daded, momed, wage76, iqscore, age76, nearc2, nearc4)

```

```{r}
#creating near college variable and dummy coding
#TZ just added the comment above to annotate what the code is doing.

schoolingsub1 <- schoolingsub |> 
  mutate(near_college = ifelse(nearc2 == "yes" & nearc4 =="yes", "near_colleges", 
                               ifelse(nearc2 == "yes" & nearc4 =="no","near_cc", 
                                      ifelse(nearc2 == "no" & nearc4 =="yes","near_4yr", "not_near_college")))) 
```

```{r}
schoolingsub1 <- schoolingsub1 %>%
  pivot_longer(
    cols = c(daded,momed),
    names_to = "parent",
    values_to = "par_education")

schoolingsub1$parent <- gsub("daded", "dad", schoolingsub1$parent)

schoolingsub1$parent <- gsub("momed", "mom", schoolingsub1$parent)
```

Demographics

```{r}


datasummary_skim(schoolingsub1 |> 
                   pivot_wider (names_from = parent,
                                values_from = par_education) |> 
                   select(-rownames),
                 histogram = FALSE)


# TZ: pretty table! maybe rename the variables to easy-to-read names; not sure if wage is weekly or montly... See revised code below:
schoolingsub2 <- schoolingsub1 %>%
  pivot_wider (names_from = parent,
               values_from = par_education) %>%
  rename("Participant Ed Level"= ed76, "Participant Weekly Wage" = wage76, "Participant IQ" = iqscore, "Participant Age" = age76, "Dad Ed Level" = dad, "Mom Ed Level" = mom)

datasummary_skim(schoolingsub2 |> 
                   select(-rownames),
                 histogram = FALSE)
```

Research question 1: What is the relationship between parent education and participant's education?

TZ: Kinda silly, but my advisor drilled into me that "people have 'relationships', data has 'relations.'" But I think that is personal preference. I would probably also put in the word "level" when talking about education because this could also mean type of education (private, public, homeschool).

```{r}
#TZ: Revised to put par_education back in and revert name of participant education to not have any spaces in variable name
  schoolingsub3 <- schoolingsub2 %>%
      rename("ed76"= "Participant Ed Level") %>%
  pivot_longer(
    cols = c("Dad Ed Level", "Mom Ed Level"),
    names_to = "parent",
    values_to = "par_education")

  ggplot(schoolingsub3, aes(par_education, ed76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + theme_minimal() +
  labs(title = "Relationship between Parent and Participant Education",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Education (yrs)",
       color = "Parent"
       )

```

Research question 2: What is the relationship between parent education and participant's wage?

TZ: same comments as RQ 1

```{r}
#TZ:  reverted wage back to original name to get rid of spaces in variable name
  schoolingsub3 <- schoolingsub3 %>%
      rename("wage76"= "Participant Weekly Wage")
  

  ggplot(schoolingsub3, aes(par_education, wage76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + theme_minimal() +
   labs(title = "Relationship between Parent Education and Participant Wages",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Wages (cents/hr)",
       color = "Parent"
       )
```

Research question 3: Is there a difference in wage depending on whether you lived near a college?

TZ: Typo on "college." This could maybe be rephrased as "Is there a difference in wages between those that lived near a college (e.g., two-year and four-year) and those that did not?

```{r}
schoolingsub1$near_college <- gsub("near_4yr", "Near 4-year College", schoolingsub1$near_college)

schoolingsub1$near_college <- gsub("near_cc", "Near Community College", schoolingsub1$near_college)

schoolingsub1$near_college <- gsub("near_colleges", "Near 4-year  and Community College", schoolingsub1$near_college)

schoolingsub1$near_college <- gsub("not_near_college", "Not Near College", schoolingsub1$near_college)

schoolingsub1 %>%   
  group_by(near_college) |> 
  summarize(avg_wage = mean(wage76, na.rm = TRUE)/100) |>
  rename("Proximity to College" = near_college, "Participant Average Wage per Hour" = avg_wage) %>%
DT::datatable()

#TZ: this is super cool! I'll have to use this one day
```
