---
title: "paper"
author: "Frances Pinkham, Eunji Kong, Elyse Calhoun, Yitong Jiang"
format: html 
embed-resources: true
execute: 
  include: true
  warning: false
bibliography: References.bib
---


```{r}
#| include: false

library(tidyverse)
library(here)
library(rio)
library(modelsummary)
library(knitr)
library(gt)
library(lme4)
```

```{r}
#| label: input-data
#| include: false

schooling <- import(here("data","Schooling.csv"))
rq1 <- "What is the relationship between parent education and participant's education?"
rq2 <- "What is the relationship between parent education and participant's wage?"
rq3 <- "Is there a difference in wage depending on whether you lived near a college?"
round <- 2
```

# Introduction

When discussing minimum wages and its economic impact, most research focuses on the impact of employment [@neumark2003]. Card [-@Card1995] suggests that growing up in an area near or around a college potentially has a direct effect on wage earnings. The assumption is made that people earn higher wages based on the amount of schooling they have completed [@dearden]. For this study, we want to determine if college proximity and the education level of parents have an impact on wage earnings.

The purpose of this study is to understand the relationship between parents' education, geographic locations (i.e., proximity to college), and participants' education and income. Specifically, given the relationships among these factors in previous literature, we sought to answer:

1.  `r rq1`
2.  `r rq2`
3.  `r rq3`

```{r}
#| label: select-variables
#| include: false

schoolingsub <- schooling %>%
  select(rownames,ed76, daded, momed, wage76, iqscore, age76, nearc2, nearc4, black)


```

```{r}
#| label: tidy-data
#| include: false

schoolingsub1 <- schoolingsub |> 
  mutate(near_college = 
           case_when(nearc2 == "yes" & nearc4 =="yes" ~ "Near 4-year and Community College",
                     nearc2 == "yes" & nearc4 =="no" ~ "Near Community College",
                     nearc2 == "no" & nearc4 =="yes"~ "Near 4-year College",
                     .default = "Not Near College"))

schoolingsub1 <- schoolingsub1 %>%
  pivot_longer(
    cols = c(daded,momed),
    names_to = "parent",
    values_to = "par_education") |> 
  mutate(parent = fct_recode(parent, 
                             "Paternal Education (Yrs)" = "daded",
                             "Maternal Education (Yrs)" = "momed"))
```

```{r}
#| label: Table1-Demographics
#| include: true
#| echo: false
# The table below lists all the data for each participants. for a demographics table, it needs to be a summary table of each participants' information 
# schoolingsub1 %>% 
#     pivot_wider (names_from = parent,
#                 values_from = par_education) %>%
#    select(-near_college) %>%
#   gt_preview(schoolingsub1, top_n = 20, bottom_n = 1) %>%
#   cols_label(ed76 = "Participant Education (Yrs)",
#              wage76 = "Participant Wage (cents/hr)",
#              iqscore = "IQ Score",
#              age76 = "Age of Participant",
#              nearc2 = "Near Community College",
#              nearc4 = "Near Four-Year College",
#              dad = "Paternal Education (Yrs)",
#              mom = "Maternal Education (Yrs)",
#              rownames = "Participant ID")

schoolingsub1 |> 
  filter(!is.na(iqscore)) |>
  pivot_wider (names_from = parent,
                values_from = par_education) |> 
  select(-rownames) |> 
  rename(
    "Participant Education (Yrs)" = ed76,
    "Participant Wage (cents/hr)" = wage76,
    "IQ Score" = iqscore,
    "Age of Participant" = age76,
    "Near Community College" = nearc2,
    "Near Four-Year College" = nearc4) |> 
  datasummary_skim(histogram = FALSE, title = "Table 1: Demograpnics")
  
sample_size <- nrow(schoolingsub)
black <- sum(schoolingsub$black == "yes")/nrow(schoolingsub)*100

avg_age <-mean(schoolingsub$age76)
age_sd <- sd(schoolingsub$age76)
min_age <-min(schoolingsub$age76)
max_age <-max(schoolingsub$age76)


avg_edu <-mean(schoolingsub$ed76)
edu_sd <-sd(schoolingsub$ed76)

avg_paredu <-mean(schoolingsub1$par_education)
paredu_sd <-sd(schoolingsub1$par_education)

avg_medu <-mean(schoolingsub$momed)
medu_sd <-sd(schoolingsub$momed)

avg_dedu <-mean(schoolingsub$daded)
dedu_sd <-sd(schoolingsub$daded)

min_iq <-min(schoolingsub$iqscore)
max_iq <-max(schoolingsub$iqscore)

min_wage <-min(schoolingsub$wage76)
max_wage <-max(schoolingsub$wage76)
```

# Method

The current study is a secondary data analysis from the National Longitudinal Survey of Young Men (NLSYM), which started in 1966 with 5525 young men between the ages of 14 and 24 in the United States. The current study looks at a subset of this data collected in 1976.

### Participants

`r sample_size` young men were included in this study. Table 1 provides demographic data of our sample. The average age of our participants in 1976 were `r round(avg_age,round)` (*SD* = `r round(age_sd,round)`) with an average of `r round(avg_edu, round)` years of education (*SD* = `r round(edu_sd, round)`). The average education of the parents of the participants were `r round(avg_paredu, round)` years of education (*SD* = `r round(paredu_sd, round)`). Interestingly, during a time when education disparity between gender was large, paternal (*M* = `r round(dedu_sd, round)`, *SD* = `r round(dedu_sd, round)`) and maternal education (*M* = `r round(medu_sd, round)`, *SD* = `r round(medu_sd, round)`) was similar in our sample.

As is true for many longitudinal studies in the mid-1960s [@Card1995], our data is not a random, representative sample of the United States population. Data was collected from neighborhoods with a high concentration of non-White residents. In the 1960s, black represented 10% of the US population [@Census1961]; however, in our study, `r round(black, round)`% of our sample were black.

### Measures

In this study, we examined participants' responses to question regarding their age, ethnicity, IQ score, parent's education (both maternal and paternal), current level of participant education, participant's current wage, and proximity to college. The age variable is a continuous variable measured in number of years (*range* = `r min_age`- `r max_age`), and the ethnicity is a binary categorical variable in which participants were asked whether they identify as black. The IQ score variable is a continuous variable (*range* = `r min_iq`- `r max_iq`). Participant and parental education were asked in number of years (continuous variable). Participant current wage was asked as cents earned per hour (*range* = `r min_wage`- `r max_wage`). Lastly, participants were asked whether they live near a 4-year college, community college, both, or neither.

### Statistical Analyses

There were no missing data in our variables, except for in the IQ scores. However, since IQ scores were included to provide demographic information, there were no need to impute data for missing data. To answer our first research question (e.g., `r rq1`) descriptively, we plotted the relationship between parent and participant's education and displayed the fitted linear line. To answer our second research question (e.g., `r rq2`) descriptively, we plotted the relationship between parent's education and participant's education and displayed the fitted linear line. Lastly, to answer our third research question (e.g., `r rq3`) descriptively, we calculated participant's wage based on their response to living proximity to college and plotted the information as a violin graph.

# Results

```{r}
#| label: Plot1
#| include: true
#| echo: false
#| message: false

schoolingsub1 |> 
  ggplot(aes(par_education, ed76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + 
  theme_minimal() +
  labs(title = "Figure 1: Relationship between Parent and Participant Education",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Education (yrs)",
       color = "Parent")

```

**Figure 1** reflects a linear regression model of participant education as a function of parental education. Although we have not performed a complete analysis on these descriptive statistics, our model may indicate a positive correlation between parental education and participant education.

```{r}
#| label: Model1
#| include: false

m1 <- lm (par_education ~ ed76, data = schoolingsub1)
coef(summary(m1))

modelsummary(list("Model" = m1),
             title = "Table 1. Model of the relationship between parents' education and participants' education",
             stars = T,
             gof_omit = "Log.Lik.|F",
             statistic = "conf.int",
             coef_rename = c("ed76" = "Participants' education")) 

```

```{r}
#| label: Plot2
#| include: true
#| echo: false
#| message: false

schoolingsub1 |> 
  ggplot(aes(par_education, wage76, color = parent)) +
  geom_smooth(method = lm, se = FALSE) + 
  theme_minimal() +
   labs(title = "Figure 2: Relationship between Parent Education and Participant Wages",
       subtitle = "From 1976 Survey",
       x = "Parental Education (yrs)",
       y = "Participant Wages (cents/hr)",
       color = "Parent"
       )

```

**Figure 2** also indicates a possible positive correlation between parental education and participant wages, shown in cents/hour. Further statistical testing would be required to verify this.

```{r}
#| label: Model2
#| include: false

m2 <- lm (par_education ~ wage76, data = schoolingsub1)
coef(summary(m2))

modelsummary(list("Model" = m2),
             title = "Table 2. Model of the relationship between parents' education and participants' wages",
             stars = T,
             gof_omit = "Log.Lik.|F",
             statistic = "conf.int",
             coef_rename = c("wage76" = "Participants' wages")) 
```

```{r}
#| label: Table3
#| include: false

prox_wage <- schoolingsub1 %>%
group_by(near_college) %>%
summarize(avg_wage = mean(wage76, na.rm = TRUE)/100) %>%
rename("Proximity to College" = near_college, 
       "Participant Average Wage per Hour" = avg_wage)%>%
  DT::datatable()
```

Those who resided near only a 4-year college had slightly lower average wages, at approximately \$5.68/hr.
Those who were not near any college had marginally higher average wages (\$5.18/hour) than those near a community college, at \$5.13 per hour.

```{r}
#| include: false

min_wage <- 2.30
near_fouryear <- 6.45
```

```{r}
#| label: Plot3
#| include: true
#| echo: false
#| message: false

ggplot(schoolingsub1, aes(x=factor(near_college), y=wage76, color=factor(near_college), fill=near_college, alpha=0.2)) +
  geom_violin() +
  stat_summary(fun = "median",
               geom = "crossbar", 
               width = 0.5,
               colour = "red") +
  ylab("Participant Wages (cents/hr)") + xlab("Participant's Proximity to a College") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```

**Figure 3** displays

```{r}
#| label: Model3
#| include: false

m3 <- lm(wage76 ~ near_college, data=schoolingsub1)
anova(m3)

```

# Discussion

# Reference
